---
title: "Homework Assignment 3"
author: "Matthew Xu"
date: "25 November 2020"
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
---

\vspace{0.25in}

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#knit from directory
setwd("/Users/MatthewXu/Desktop/PSTAT131")

library(readr)

knitr::opts_knit$set(root.dir="/Users/MatthewXu/Desktop/PSTAT131")
```

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
library(tidyverse) 
library(ROCR) 
library(tree) 
library(maptree) 
library(class) 
library(lattice) 
library(ggridges) 
library(superheat)
```

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
drug_use <- read_csv('drug.csv',
col_names = c('ID','Age','Gender','Education','Country','Ethnicity',
                                 'Nscore','Escore','Oscore','Ascore','Cscore','Impulsive',
                                'SS','Alcohol','Amphet','Amyl','Benzos','Caff','Cannabis',
              'Choc','Coke','Crack','Ecstasy','Heroin','Ketamine',
              'Legalh','LSD','Meth','Mushrooms','Nicotine','Semer','VSA'))
```

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
drug_use <- drug_use %>% mutate_at(as.ordered, .vars=vars(Alcohol:VSA)) 
drug_use <- drug_use %>%
  mutate(Gender = factor(Gender, labels=c("Male", "Female"))) %>% 
  mutate(Ethnicity = factor(Ethnicity, labels=c("Black", "Asian", "White",
"Mixed:White/Black", "Other", "Mixed:White/Asian", "Mixed:Black/Asian"))) %>%
  mutate(Country = factor(Country, labels=c("Australia", "Canada", "New Zealand", "Other", "Ireland", "UK", "USA")))
```

# 1. "PROBLEM NUMBER ONE"

a. "PROBLEM NUMBER ONE, PART A"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#mutate to factor yes or no if >= CL3
drug_use <- drug_use %>%
  mutate(recent_cannabis_use= factor(ifelse(drug_use$Cannabis >= "CL3", "Yes", "No"),  labels = c("No", "Yes")))

#test to see if factor
class(drug_use$recent_cannabis_use)
```

b. "PROBLEM NUMBER ONE, PART B"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
drug_use_subset <- drug_use %>% select(Age:SS, recent_cannabis_use)

#split into train.test sets of train set size = 1500
set.seed(1, sample.kind = "Rounding")
train.indices = sample(1:nrow(drug_use_subset), 1500)
drug_use_train=drug_use_subset[train.indices,]
drug_use_test=drug_use_subset[-train.indices,]

#dimensions of train should be 1500, test 385 (1835 - 1500)
dim(drug_use_train)
dim(drug_use_test)

```

c. "PROBLEM NUMBER ONE, PART C"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#fitting logitisc regression to train data
glm.fit <-glm(drug_use_train$recent_cannabis_use~., data=drug_use_train, family= binomial)
summary(glm.fit)
```

# 2. "PROBLEM NUMBER TWO"

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#tree control
tree_parameters = tree.control(nobs=nrow(drug_use_train), minsize=10, mindev=1e-3)
```

a. "PROBLEM NUMBER TWO, PART A"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}

nfold = 10
folds = seq.int(nrow(drug_use_train)) %>% ## sequential obs ids
cut(breaks = nfold, labels=FALSE) %>% ## sequential fold ids
sample ## random fold ids

drugtree <- tree(drug_use_train$recent_cannabis_use~., drug_use_train, control = tree_parameters)

set.seed(1, sample.kind = "Rounding")
# K-Fold cross validation
#rand is number of cases, the fold partioning
cv = cv.tree(drugtree, rand = folds,  K=10, FUN=prune.misclass)

#find best tree size
#there are identical deviations for different sizes 
#min of cv$size from the minimum positions of the minimum of cv$dev
cv$size
cv$dev

best_size = min(cv$size[which(cv$dev == min(cv$dev))])
best_size
```
Best size of the pruned tree is of 8 internal nodes.

b. "PROBLEM NUMBER TWO, PART B"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#prune tree to best size
drugtree.pruned = prune.tree(drugtree, best=best_size)

# Plot the tree
draw.tree(drugtree.pruned, nodeinfo=TRUE) 
title("Classification Tree Built on Training Set")
```
The first variable split in this decision tree is Country.

c. "PROBLEM NUMBER TWO, PART C"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
# Predict on test set
tree.pred = predict(drugtree.pruned, drug_use_test, type="class") 

#confusion matrix for truths/falses
test_labels = drug_use_test$recent_cannabis_use
test.table = table(tree.pred, test_labels)
test.table

#TPR = TP/(TP+FN)
#FPR = FP/(FP+TN)
TPR = test.table[2,2]/(test.table[2,2] + test.table[1,2])
TPR
FPR = test.table[2,1]/(test.table[2,1] + test.table[1,1])
FPR
```

# 3. "PROBLEM NUMBER THREE"

a. "PROBLEM NUMBER THREE, PART A"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#prob training for logisitic regression 
#fit model to test data
prob.training1 = predict(glm.fit, drug_use_test, type="response")
#prob training for decision tree 
prob.training2 = predict(drugtree.pruned, drug_use_test, type = "vector")

# First arument is the prob.training, second is true labels
LOGpred = prediction(prob.training1, drug_use_test$recent_cannabis_use)
DTpred = prediction(prob.training2[,2], drug_use_test$recent_cannabis_use)

# We want TPR on the y axis and FPR on the x axis
perf1 = performance(LOGpred, measure="tpr", x.measure="fpr")
perf2 = performance(DTpred, measure="tpr", x.measure="fpr")

#plot both ROC curves on same plot
plot(perf1, col=2, lwd=3, main="ROC curve") 
par(new = TRUE)
plot(perf2, col=3, lwd=3, main="ROC curve")
abline(0,1)

```
b. "PROBLEM NUMBER THREE, PART B"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#Calculate AUC log
auc1 = performance(LOGpred, "auc")@y.values 
auc1

#Calculate AUC Descision Tree
auc2 = performance(DTpred, "auc")@y.values 
auc2
```
The logistic Regression model has the better AUC value.

# 4. "PROBLEM NUMBER FOUR"

```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
leukemia_data <- read_csv("leukemia_data.csv")
```

a. "PROBLEM NUMBER FOUR, PART A"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#change variable Type to factor
leukemia_data <- leukemia_data %>%
  mutate_at(vars('Type'), as.factor)

#check to see if factor
class(leukemia_data$Type)

#frequency table for each leukemia subtype
table(leukemia_data$Type)
```
The BCR-ABL leukemia subtype occurs the least.

b. "PROBLEM NUMBER FOUR, PART B"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#using prcomp to calculate pve and cumulative pve
leukemia_pca <- prcomp(leukemia_data[, -c(1)], scale = TRUE, center = TRUE)
sdev <- leukemia_pca$sdev
pve <- sdev^2/sum(sdev^2)
cumulative_pve <- cumsum(pve)

## This will put the next two plots side by side 
par(mfrow=c(1, 2))

## Plot proportion of variance explained \
plot(pve, type="l", lwd=3) 
plot(cumulative_pve, type="l", lwd=3)

```

c. "PROBLEM NUMBER FOUR, PART C"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#load colors
require(graphics)
rainbow_colors <- rainbow(7)
plot_colors <- rainbow_colors[leukemia_data$Type]

new_coords <- leukemia_pca$x[,1:2]
plot(new_coords, cex = 0.5)
#from piazza do not need -new_coords
text(new_coords, labels=leukemia_data$Type, col = plot_colors, cex = 0.6)

sorted_PC1 <- sort(abs(leukemia_pca$x[,1]), decreasing = TRUE)
head(sorted_PC1, n = 6)
```
Along the PC1 axis, the group that appears to be the most seperated from the others is the T-All gene. The six genes from PC1 that have the largest weighted have weights of 52.06836, 51.75384, 49.06527, 48.16035, 48.09339, 47.96769.

f. "PROBLEM NUMBER FOUR, PART F"
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
library(dendextend)

#subset data to these 3 leukemia subtypes
leukemia_subset = filter(leukemia_data, Type == "T-ALL" | Type == "TEL-AML1" | Type == "Hyperdip50")

#eucledian distance with compltete linkage
dis = dist(leukemia_subset, method="euclidean")
set.seed(213)
leukemia.hc = hclust(dis, method="complete")

## dendrogram: branches colored by 3 groups 
dend1 = as.dendrogram(leukemia.hc)
# color branches and labels by 3 clusters 
dend1 = color_branches(dend1, k=3)
dend1 = color_labels(dend1, k=3)
# change label size
dend1 = set(dend1, "labels_cex", 0.3)

# add true labels to observations
dend1 = set_labels(dend1, labels=leukemia_subset$Type[order.dendrogram(dend1)]) # plot the dendrogram
plot(dend1, horiz=T, main = "Dendrogram colored by three clusters")

## dendrogram: branches colored by 5 groups 
dend2 = as.dendrogram(leukemia.hc)
# color branches and labels by 5 clusters 
dend2 = color_branches(dend2, k=5)
dend2 = color_labels(dend2, k=5)
# change label size
dend2 = set(dend2, "labels_cex", 0.3)

# add true labels to observations
dend2 = set_labels(dend2, labels=leukemia_subset$Type[order.dendrogram(dend2)]) # plot the dendrogram
plot(dend2, horiz=T, main = "Dendrogram colored by five clusters")
```