---
title: "PSTAT 105 HW 7"
author: "Matthew Xu (5752811)"
date: "7 March 2021"
output:
  pdf_document: default
  html_document: default
  word_document: default
fontsize: 11pt
geometry: margin=1in
header-includes:
- \usepackage{amsmath}
---

\vspace{0.25in}
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#libraries
library(tidyverse)

#scan in signal data
signal<- read_delim("andro2.txt", " ", col_names = T)
```


# Please complete these questions and submit the answers on GauchoSpace. Your calculations should be done in R, and your results should be typed up to include R input and output as well as clear explanations for your answers

1. The file andro2.txt contains the record of biological signal taken on a mouse subject. We are going
to model this data as noisy measurements around an unknown mean function.

(a) Esimate the value of the mean function at time 200 by first averaging the 35 points closest to that
time, and then average the 105 points closest to time 200. Give 95% confidence intervals for each
of these estimates using a t statistic and estimating the variance from the data used in the mean.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
n=35
signal$dist_200 = abs(signal$Time - 200)
signal_35 = signal %>% arrange(dist_200) %>% top_n(-35, dist_200)
x_bar = mean(signal_35$Signal)

var = var(signal_35$Signal)
var

t_stat = qt(0.975, n-1)
t_stat

s = sum((signal_35$Signal-x_bar)^2)
#including zero includes the mean and lower/upper bounds 95% CI
CI = x_bar + c(-t_stat, 0, t_stat)*sqrt(s/n)
CI
```
95% of the confidence can be explained between -0.18125269 and 0.22639554 with the true mean being within those bounds. THe variance of the mean is about 0.01035496. The test statistic is about 2.032245.

(b) In measuring the performance of a nonparametric regression estimator, we are looking at the
variance and the bias of the estimate. Which of these two is not accounted for in your 95%
confidence intervals from part (a)? Explain.

If the non paramaeric funtion is not normal, in reference to the t test statistic, the variance is usually underestimated. Confidence intervals require variance to be calculated, therefore accounting variance. Bias is not considered when calcuating confidence interval in performance of a nonparamteric regression estimator.

(c) Plot the data with time as the x axis and signal as the y axis. Calculate the Nadaraya-Watson
kernel estimate using a Gaussian kernel, and plot the resulting estimate of a smooth mean function.
Use a bandwdith that you think works well.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
k_est = ksmooth(signal$Time, signal$Signal, kernel = "normal", bandwidth = 1)
k_est_d = data.frame(k_est$x, k_est$y)

x = ggplot(signal, aes(x = Time, y = Signal)) + geom_point()
x + geom_line(data = k_est_d, aes(x= k_est.x, y = k_est.y), col = 'blue')
```

(d) Make a new scatter plot and then add a local linear regression estimate using the locpoly function
from the KernSmooth library. Choose a new bandwidth that fits the data well.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
library(KernSmooth)
loc = locpoly(signal$Time, signal$Signal, kernel = 'normal', bandwidth = 0.5)
loc_d = data.frame(loc$x, loc$y)

x + geom_line(data = loc_d, aes(x= loc.x, y = loc.y), col = 'lightblue')
```

(e) The scientists were looking at these measurements for evidence of a sequence of peaks in the levels
happening regularly over time. Use your estimates of the mean functions to locate the peaks in
the data stream, and calculate the average amount of time between the peaks.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
library(pracma)
loc_peaks = findpeaks(loc$y, npeaks=10, threshold = 0.2, sortstr=TRUE)
l_est_peaks = findpeaks(k_est$y, npeaks=10, threshold = 0.2, sortstr=TRUE)

loc_peaks
l_est_peaks

time_between_peaks1 = loc_peaks[1,2] - loc_peaks[2,2]
time_between_peaks2 = loc_peaks[1,3] - loc_peaks[2,3]
time_between_peaks3 = loc_peaks[1,4] - loc_peaks[2,4]
time_between_peaks4 = l_est_peaks[1,2] - l_est_peaks[2,2]
time_between_peaks5 = l_est_peaks[1,3] - l_est_peaks[2,3]
time_between_peaks6 = l_est_peaks[1,4] - l_est_peaks[2,4]

time_between_peaks1
time_between_peaks2
time_between_peaks3
time_between_peaks4
time_between_peaks5
time_between_peaks6
```

A peak is define here as the highest point of the distrbution using a spesfic kernel for a given amount of time. The peaks here are the high points on the plot after a set of low points. Certain time intervals have mutliple similar high points such as around 600 for the locpoly plot. We can access them thorugh the time intervals for where they occur in. We can use the findpeaks function to find the peaks at certain time points given the number of peaks and the threshold for finding the peaks. The time between peaks in loc is about 130. The time between peaks in l_est is about 340.

2. Use a simulation of one million random trials to approximate the probability for an exponential random
variable X with E(X) = 1 of the event A = {cos(X) > 0.7}. Calculate a margin of error for this
approximation by using 95% confidence interval from a binomial experiment.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
B = 100
n = 1000000
p = NULL
samples = list()

for (i in 1:B){
  samples[[i]] = rexp(n,1)
  p[i] = length(which(cos(samples[[i]]) > 0.7))/n
}

mean(p) + c(-1.96, 0, 1.96)*sqrt(mean(p)*(1-mean(p))/n*B)
```

3. We want to calculate a critical value for a Goodness of Fit test based on the Kolmogorov–Smirnov test
for an exponential distribution.

(a) Generate 1000 samples where each consists of 50 independent exponential random variables with
mean 1. Estimate the mean of each sample. Draw a histogram of the means.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
n2 = 50
samples2 = list()
mu = NULL

for (i in 1:1000){
  samples2[[i]] = rexp(n2,1)
  mu[i] = mean(samples2[[i]])
}

g = ggplot(data = data.frame(mu), aes(x=mu)) + geom_histogram(bins = 40, fill = 'red', color = 'black')
g
```

(b) Perform a KS test on each sample against the null hypothesis that they are from an exponential
random variable with a mean that matches the mean of the data set. Draw a histogram of the
1000 values of D.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
D = NULL

for (i in 1:1000){
  D[i] = ks.test(samples2[[i]], "pexp", mu[i])$statistic
}

h = ggplot(data = data.frame(D), aes(x = D)) + geom_histogram(bins = 35, fill = "green", color = 'black')
h
```

(c) From the simulated values of D, find an critical value c such that only 5% of the time will the
test statistic D exceed that critical value.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
quantile(D,0.95)
```

(d) Write up R code that will perform multiple batches of 1000 samples each consisting of 50 independent draws. Decide how many batches your computer can run in two hours, and run the program.
Use the results to generate a more accurate estimate of c.
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
K = 5000
c = NULL
M = matrix(nrow = K, ncol = 1000)
a = Sys.time()
for (i in 1:K){
  samples3 = list()
  for (j in 1:1000){
    samples3[[j]] = rexp(50,1)
    M[i,j] = ks.test(samples3[[j]], 'pexp', mean(samples3[[j]]))$statistic
  }
  c[i] = quantile(M[i,], 0.95)
}
b = Sys.time()
a-b
mean(c)
```

(e) Calculate the critical value that is suggested in Table 1.4 of Stephens(1974) paper. Compare this
to the results from your simulations. Do you think this value is significantly different from your
simulation results?
```{r tidy = TRUE, message = FALSE, warning = FALSE, error = FALSE}
#n value in D.star is value in rexp not simulation number
(D.star <- 0.895/(sqrt(50) -0.01 +0.85/sqrt(50)))
t.test(c, mu = D.star)
```
No, they are quite similar numerically with the 95% percentile using D values at about 0.2648041, the simulated 95% of mean at about 0.2668, and the calculated mean using Stephens paper at about 0.2671479. They are very close to each other.

